---
title: "Art Price Index"
author: "Laurie Binge"
date: "30 November 2015"
output: 
    pdf_document:
        fig_caption: yes
        number_sections: true
bibliography: References.bib
---

#Introduction
Contemporary African art, long seen as a niche market, has experienced a surge in popularity over the last few decades. The South African art market in particular has received a lot of attention and has grown markedly over the last two decades, both in terms of the number of transactions and total turnover [@Fedderke2014]. Artworks by South African artists have reached record prices at international and local auctions, both for the country's "masters" - including Irma Stern, Walter Battiss, and JH Pierneef - and contemporary artists like William Kentridge [@Naidoo2013]. For example, in 2011 Irma Stern's "Arab Priest" set a world record hammer price of £2.7 million at auction (Bonhams), while "Two Arabs" sold for R19 million, a record for a South African auction house (Strauss & Co). 

The increase in interest in South African art, both locally and abroad, has sparked a vibrant market for investors [@Naidoo2013]. This increase in the popularity of art, partly as an investment vehicle, is commensurate with international trends, where fine art has become an important asset class in its own right. In 2010 the Wall Street Journal reported that around 6% of total wealth was held in so-called passion investments, which include art, wine, antiques and jewellery. Of all these luxury goods, art is the most likely to be acquired for its potential appreciation in value [@Capgemini2010]. Passion investments, and art in particular, are interesting examples of alternative assets, as they are durable goods with investment as well as consumption characteristics [@Renneboog2014].

In times of economic uncertainty there is often an increase in demand for physical assets: as these have limited supply they are often considered relatively safe in times of financial turmoil [@Warwick-Ching2013]. In addition, the demand for alternative assets is supported by their imperfect correlation with the stock market, which is thought to aid portfolio diversification. Alternative assets are also used as collateral for loans, or to take advantage of slacker regulatory and tax provisions. 

To date there has been little research on the South African art market and particularly trends in art prices. It is important to analyse price movements over time in order to understand the dynamics of the market and to answer some question around the development of this market. However, accurate valuation of real alternative assets like art can be difficult. These assets are heterogeneous and often involve large transaction costs for both buyers and sellers. They are less liquid than traditional assets and have a low transaction frequency, which makes it difficult to measure the state of the overall market, as only a small part of the overall market is traded at any given time. 

This paper will attempt to estimate an accurate price index for South African art. The price indices are intended to be a summary of overall price movements in the art market. The indices are then be used to try to answer the questions of whether there was a large increase in prices in the run-up to the Great Recession and whether there is evidence for the presence of a bubble in the market, as is often claimed? Section 2 provides an outline of the methodologies applied in the literature and provides a brief literature review. Section 3 looks at the available data for South Africa. Section 4 reports the results from a number of potential estimation methods. Section 5 evaluates these results and compares the indices to international art price indices. Section 7 introduces the bubble detection methodology and briefly looks at the literature. Section 8 reports the results of the bubble detection evidence. Section 9 concludes.

```{r cleaning, echo=FALSE, results='hide', message=FALSE, warning=FALSE, cache = TRUE}
##=====================##
## READING IN THE DATA ##
##=====================##
suppressMessages(library(zoo))
suppressMessages(library(ggplot2))
suppressMessages(library(plyr))
suppressMessages(library(dplyr))
suppressMessages(library(reshape2))
suppressMessages(library(stargazer))
suppressMessages(library(micEcon))
suppressMessages(library(quantreg))
suppressMessages(library(McSpatial))
suppressMessages(library(quantmod))
suppressMessages(library(xtable))

setwd("C:\\Users\\Laurie\\OneDrive\\Documents\\BING\\PhD Proposal Readings\\Art Price Index\\R Code")
artdata <- read.csv("Auction database.csv", header=TRUE, sep=";",na.strings = "", skipNul = TRUE, 
                    colClasses=c("character","numeric","numeric","numeric","numeric","factor","factor","factor","character",
                                 "factor","factor","factor","character","factor","factor","factor","numeric","character",
                                 "numeric","numeric","numeric","numeric","numeric","numeric"))
##===================##
## CLEANING THE DATA ##
##===================##

artdata$date <- as.Date(artdata$date)
artdata$med_code <- factor(artdata$med_code, labels=c("Drawing", "Watercolour", "Oil", "Acrylic", "Print/Woodcut",
                                                      "Mixed Media","Sculpture","Photography", "Other"))
artdata$ah_code <- factor(artdata$ah_code, labels=c("5th Avenue","Ashbeys","Bernardi","Bonhams","Russell Kaplan",
                                                    "Stephan Welz","Strauss","Christies"))
artdata$timedummy <- factor(as.yearqtr(artdata$date, "%Y-%m-%d"))
artdata$lnprice <- log(artdata$price)
artdata$lnarea <- log(artdata$area)
artdata$lnarea2 <- artdata$lnarea*artdata$lnarea
artdata$lnsculpt_area <- ifelse(artdata$med_code=="Sculpture", artdata$lnarea, 0)
artdata$counter <- as.numeric(artdata$timedummy)

##----------------------
##Rank Artists by Volume
##----------------------
#Rank by Total Volume (all)
rankings <- count(artdata, artist)
rankings$rank_all <- dense_rank(desc(rankings$n))    #rank by density, with no gaps (ties = equal)
rankings$rank_total <- row_number(desc(rankings$n))  #equivalent to rank(ties.method = "first")
rankings$n <- NULL

artdata <- merge(artdata, rankings, by.x="artist", by.y="artist",all.x=TRUE)
```

#Estimation methodologies
An accurate measure is a prerequisite to analysing the art market to try and examine whether there was evidence of a bubble. The aim of this section is to establish a range of measures to answer the questions of .

The construction of price indices for real alternative asset markets is challenging for at least two reasons [@Jiang2014]. Firstly, the low frequency of trading means that only a subset of the market is traded at a given time, while the prices of non-transacted assets are unobservable. Secondly, the heterogeneity of individually unique assets means that the quality of assets sold is not constant over time. Thus, the composition of assets sold will generally differ between periods, making it difficult to compare prices over time [@Hansen2009]. Constructing an index for individually unique assets, like art, therefore requires a different approach than is used for indices of stocks, bonds and commodities. Four broad measurement techniques have been used to construct these indices [@Eurostat2013]:

a)	Naïve or central tendency methods
b)	Hedonic regressions
c)	Repeat sales regressions
d)	Hybrid models

The following sections provide a brief introduction to these methodologies. The literature does not provide an a priori indication of the most appropriate method and, in practice, the data dictates the choice.

##Central Tendency or naïve methods
The simplest way to construct a price index is to calculate a measure of central tendency from the distribution of prices. As price distributions are generally skewed, the median is often preferred to the mean. These average measures have the advantage that they are simple and easy to construct and do not require detailed data. 

However, an index based on average prices does not account for the difficulties mentioned above. For assets like artworks, naïve indices may therefore be more dependent on the mix of objects that come to market, than changes in the underlying market. For instance, if there is an increase in the share of higher quality assets, an average measure will show an increase in price, even if the prices in the market did not change [@Hansen2009]. Hence, such a measure may not be representative of the price movements of all the assets in the market. If there is a correlation between turning points in asset price cycles and compositional and quality changes, then an average could be especially inaccurate [@Eurostat2013].

An improvement can be made by stratification of the data. Stratified measures control for variations in prices across different types of assets by separating the sample into subgroups according to individual characteristics such as artist and medium.[^1] Stratified measures are currently used by ABSA, FNB and Standard Bank, for instance, to construct property price indices for South Africa. However, scholarly work rarely employs central tendency indices. The repeat sales and the hedonic regression methods have dominated in the international literature. 

[^1]: Ek het weergawes van hierdie indekse bereken, so ons kan hulle insluit as dit nodig is. The Fisher ideal index is often the recommended index formula, as it can be justified from several different perspectives (Eurostat, 2013). It is the geometric mean of the Laspeyres and Paasche indices. The Laspeyres index holds the quantity weights fixed in the base period, while the Paasche index holds the quantity weights fixed at the comparison period.

##Hedonic

$$\ln P_{it} =\alpha+\sum_{j=1}^z\beta_jX_{ij}+\sum_{t=0}^\tau\gamma_tD_{it}+\epsilon_{it}$$

###Reputaion variable

$$Index = \frac{\prod_{i=1}^n(P_{i,t+1})^\frac{1}{n}}{\prod_{i=1}^m(P_{i,t})^\frac{1}{m}}/\text{hedonic adjustment} $$

$$\text{hedonic adjustment} = \exp \left[\sum_{j=1}^z\beta_j(\sum_{i=0}^n \frac{X_{ij,t+1}}{n}- \sum_{i=1}^m \frac{X_{ij,t}}{m})\right]$$

$$\text{Artist reputation index} = \frac{\prod_{i=1}^n(P_{i,y})^\frac{1}{n}/\prod_{i=1}^m(P_{i,0})^\frac{1}{m}}{\exp \left[\sum_{j=1}^z\beta_j(\sum_{i=0}^n \frac{X_{ij,y}}{n}- \sum_{i=1}^m \frac{X_{ij,0}}{m})\right]} $$


##Repeat Sales

$$\ln\frac{P_{t+1}}{P_t} =\sum_{i=1}^t\gamma_id_{i}+\epsilon_{i}$$


##Hybrid Models: pseudo-Repeat Sales

$$\ln P_{bsj} - \ln P_{arj}=\sum_{k=1}^K\beta_k(X_{bsjk}-X_{arjk})+\sum_{t=0}^\tau\gamma_td_{t}+\epsilon_{srabj}$$
#Lit Review

## South African literature


#The South African Art Auction Data
Daar was baie groei.

```{r figure1, echo=FALSE, cache = TRUE, fig.height=4, fig.width=7, fig.cap="Turnover (sum of hammer prices) by auction house (2000-2015YTD)"}
artplot <- aggregate(artdata$hammer_price, by=list(artdata$year,artdata$ah_code), FUN = sum, na.rm=TRUE)
g <- ggplot(artplot, aes(x=Group.1, y=x,fill=Group.2))
g <- g + geom_bar(stat="identity")
g <- g + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
g <- g + scale_fill_discrete(name="Auction House")
g <- g + ylab("Turnover (Sum of Hammer Price)")
g <- g + xlab("Date")
g
```

En die pryse het gestyg.

```{r figure2, echo=FALSE, cache = TRUE, fig.height=4, fig.width=6, fig.cap="Median hammer prices and total sales (lots) at auction (2000-2015YTD)"}
artplot1 <- aggregate(artdata$hammer_price, by=list(artdata$year), length)
artplot2 <- aggregate(artdata$hammer_price, by=list(artdata$year), FUN = median, na.rm=TRUE)
artplot <- merge(artplot1, artplot2, by="Group.1",all.x=TRUE)
names(artplot) <- c("Date","Total Sales","Median Price")
artplot <- melt(artplot, id="Date") 
g <- ggplot(artplot, aes(x=Date,value,colour=variable,fill=variable))
g <- g + geom_bar(subset=.(variable=="Total Sales"),stat="identity")
g <- g + geom_line(subset=.(variable=="Median Price"),size=1)
g <- g + theme(legend.position="bottom") + theme(legend.title=element_blank())
g
```

Summary van die pryse

```{r table1, echo=FALSE, results='asis', message=FALSE, cache = TRUE}
summaryfunction <- function(x) {
    if( is.numeric(x)!=TRUE) {stop("Supplied X is not numeric")}
    mysummary = data.frame("Min." =as.numeric(min(x,na.rm=TRUE)),"1st Qu." = quantile(x,na.rm=TRUE)[2],
                           "Median" = median(x,na.rm=TRUE),"Mean" = mean(x,na.rm=TRUE),"3rd Qu." = quantile(x,na.rm=TRUE)[4],
                           "Max." = max(x,na.rm=TRUE),row.names="")
    names(mysummary) = c("Min.","1st Qu.","Median","Mean","3rd Qu.","Max.")
    return(mysummary)
}
xt <- xtable(summaryfunction(artdata$hammer_price), caption="Descriptive statistics of auction hammer prices")
print(xt, "latex",comment=FALSE)
```


#Results
##Hedonic


```{r hedonicrep, echo=FALSE, cache = TRUE}
##------------------------------------------##
##---ARTIST REPUTATION VARIABLE (Kraussl)---##
##------------------------------------------##
#First step: Estimate equation 1 on a sub-sample of artists to obtain betaj coefficients
modeldata <- subset(artdata, artdata$rank_all<101)
list_expl_vars <- c("lnarea","ah_code","med_code","lnsculpt_area","dum_signed", "dum_dated",  
                    "nr_works","artist","timedummy")
expl_vars <- as.formula(paste("lnprice~",paste(list_expl_vars,collapse="+")))
model_100 <- lm(expl_vars, data=modeldata)

#Second step: betaj coefficients are plugged into equation for every artist pair (base & another) 
rep <- list()
rep[[1]] <- 1
for(i in 2:(max(artdata$rank_total))) {
    list_vars <- c(list_expl_vars,"price")
    
    #geometric mean of paintings by artist y
    y <- subset(artdata[,list_vars], artdata$rank_total==1)
    y <- y[!rowSums(is.na(y)), ]
    py <-  exp(mean(log(y$price)))  
    ym1 <- subset(artdata[,list_vars], artdata$rank_total==i)
    ym1 <- ym1[!rowSums(is.na(ym1)), ]
    pym1 <-  exp(mean(log(ym1$price)))
    sbx <- 0
    
    #average of characteristics time implicit attribute price
    xy <- mean(y$lnarea)
    xym1 <- mean(ym1$lnarea)   
    b <- summary(model_100)$coefficients[grepl("lnarea", rownames(summary(model_100)$coefficients)),1]
    bx <- b*(xym1-xy)   
    sbx <- sbx + bx
    
    xy <- mean(y$lnsculpt_area)
    xym1 <- mean(ym1$lnsculpt_area)   
    b <- summary(model_100)$coefficients[grepl("lnsculpt_area", rownames(summary(model_100)$coefficients)),1]
    bx <- b*(xym1-xy)   
    sbx <- sbx + bx
    
    xy <- mean(y$nr_works)
    xym1 <- mean(ym1$nr_works)   
    b <- summary(model_100)$coefficients[grepl("nr_works", rownames(summary(model_100)$coefficients)),1]
    bx <- b*(xym1-xy)   
    sbx <- sbx + bx
    
    xy <- mean(as.numeric(y$dum_signed)-1)
    xym1 <- mean(as.numeric(ym1$dum_signed)-1)   
    b <- summary(model_100)$coefficients[grepl("dum_signed", rownames(summary(model_100)$coefficients)),1]
    bx <- b*(xym1-xy)   
    sbx <- sbx + bx
    
    xy <- mean(as.numeric(y$dum_dated)-1)
    xym1 <- mean(as.numeric(ym1$dum_dated)-1)   
    b <- summary(model_100)$coefficients[grepl("dum_dated", rownames(summary(model_100)$coefficients)),1]
    bx <- b*(xym1-xy)   
    sbx <- sbx + bx
    
    auc_house <- c("Ashbeys","Bernardi","Bonhams","Russell Kaplan","Stephan Welz","Strauss","Christies")  
    for(j in auc_house) {
        xy <- mean(as.numeric(y$ah_code==j))
        xym1 <- mean(as.numeric(ym1$ah_code==j))  
        b <- summary(model_100)$coefficients[grepl(j, rownames(summary(model_100)$coefficients)),1]
        bx <- b*(xym1-xy)   
        sbx <- sbx + bx
    }
    
    medium <- c("Watercolour","Oil","Acrylic","Print/Woodcut","Mixed Media","Sculpture","Photography","Other")  
    for(k in medium) {
        xy <- mean(as.numeric(y$med_code==k))
        xym1 <- mean(as.numeric(ym1$med_code==k))   
        b <- summary(model_100)$coefficients[grepl(k, rownames(summary(model_100)$coefficients)),1]
        bx <- b*(xym1-xy)   
        sbx <- sbx + bx
    }
    rep[i] <- (pym1/py)/exp(sbx)
}

for(i in 1:(max(artdata$rank_total))) { 
    artdata$reputation[(artdata$rank_total==i)] <- rep[i]
}

artdata$reputation <- as.numeric(unlist(artdata$reputation))
artdata$lnrep <- log(artdata$reputation)

#The result: index of average price per artist adjusted for quality, relative to the base artist 
#It can replace the artist dummies as a continuous variable in a second regression of equation 1 
list_expl_vars <- c("lnarea","ah_code","med_code","lnsculpt_area","dum_signed","dum_dated",  
                    "nr_works","lnrep","timedummy")
source("full_model.R")
rep_results <- full_model(artdata,list_expl_vars)

modeldata <- artdata
model_all <- lm(expl_vars, data=modeldata)
stargazer(model_all, omit=c("timedummy"),omit.labels = "Quarterly dummies",type = "latex")

source("overlap1y_model.R")
rep_overlap1 <- overlap1y_model(artdata,list_expl_vars)

source("overlap2y_model.R")
rep_overlap2 <- overlap2y_model(artdata,list_expl_vars)

source("rolling_model.R")
rep_rolling <- rolling_model(artdata,list_expl_vars)

hedonic_indices <- rep_rolling[,c(1,2)]
colnames(hedonic_indices) <- c("Date","Rep_Full")
hedonic_indices <- cbind(hedonic_indices,RepAdjacent_1y=rep_overlap1[,19])
hedonic_indices <- cbind(hedonic_indices,RepAdjacent_2y=rep_overlap2[,11])
hedonic_indices <- cbind(hedonic_indices,RepRolling=rep_rolling[,15])
```

```{r figure3, echo=FALSE, cache = TRUE, fig.height=4, fig.width=6, fig.cap="Hedonic South African art price indices (2000Q1=100)"}
index_plot <- melt(hedonic_indices, id="Date")  # convert to long format
g <- ggplot(data=index_plot,aes(x=Date, y=value, group=variable, colour=variable)) 
g <- g + geom_point(size = 3) 
g <- g + geom_line()
g <- g + ylab("Index")
g <- g + xlab("")
g <- g + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
g <- g + theme(legend.position="bottom") + theme(legend.title=element_blank())
g
```


##Hybrid Models: pseudo-Repeat Sales

```{r repeatsales, echo=FALSE, cache = TRUE}

allDup <- function(value) {
    duplicated(value) | duplicated(value, fromLast = TRUE)
}
rsartdata <- artdata[allDup(artdata[,c("artist","title","med_code","area","dum_signed","dum_dated")]),]
rsartdata <- transform(rsartdata, id = as.numeric(interaction(artist,factor(title),med_code,factor(area),factor(dum_signed),
                                                              factor(dum_dated), drop=TRUE)))

repdata <- repsaledata(rsartdata$lnprice,rsartdata$counter,rsartdata$id)
repeatsales <- repsale(repdata$price0,repdata$time0,repdata$price1,repdata$time1,mergefirst=2,
                       graph=TRUE,graph.conf=TRUE,conf=.95)
repeatsales_index <- exp(as.data.frame(repeatsales$pindex))*100
```

There are only 
```{r, echo=FALSE} 
nrow(repdata)
```
repeat sales in the sample.

So we turn to pseudo Repeat Sales.
```{r psRS, echo=FALSE, cache = TRUE}
##------------------------------------------------------------------
##--------------- Pseudo Repeat Sales ------------------------------
##------------------------------------------------------------------
#The distance metric between any two sales (across the intervening time period) is the 
#absolute value of the difference between the two predicted hedonic log values. 
#The threshold for this distance metric can be customised. 
#At one extreme, one can choose to select only one pair with the smallest value of the distance metric. 

#match per artist by hedonic function
list_expl_vars <- c("lnarea","ah_code","med_code","lnsculpt_area","dum_signed","dum_dated",  
                    "nr_works","timedummy")
expl_vars <- as.formula(paste("lnprice~",paste(list_expl_vars,collapse="+")))

ps.RS <- function(threshold) {
    teller <-0
    repdata <- data.frame()
    rartdata <- subset(artdata,artdata$rank_all<max(artdata$rank_all,na.rm=TRUE))
    keep <- c("lnprice","artist","title","lnarea","ah_code","med_code","lnsculpt_area","dum_signed","dum_dated",  
              "nr_works","timedummy","counter")
    rartdata <- rartdata[,names(rartdata) %in% keep]
    rartdata <- rartdata[complete.cases(rartdata),]  

    ## Rank total again for new sample
    rankings <- count(rartdata, artist)
    rankings$rank_total <- row_number(desc(rankings$n))
    rartdata <- merge(rartdata, rankings, by.x="artist", by.y="artist",all.x=TRUE)
    
    for(k in 1:max(rartdata$rank_total)) { #max(rartdata$rank_total)
        modeldata <- subset(rartdata, rartdata$rank_total==k)
        modeldata$med_code <- factor(modeldata$med_code)
        modeldata$ah_code <- factor(modeldata$ah_code)
        modeldata$dum_signed <- factor(modeldata$dum_signed)
        modeldata$dum_dated <- factor(modeldata$dum_dated)
        modeldata$timedummy <- factor(modeldata$timedummy)
    
    #modeldata <- cbind(modeldata,model.matrix(~modeldata$med_code))
        if(length(levels(modeldata$med_code))==1)   { modeldata$med_code <- as.numeric(0) }
        if(length(levels(modeldata$ah_code))==1)    { modeldata$ah_code <- as.numeric(0) }
        if(length(levels(modeldata$dum_signed))==1) { modeldata$dum_signed <- as.numeric(0) }
        if(length(levels(modeldata$dum_dated))==1)  { modeldata$dum_dated <- as.numeric(0)   }
    
        if(length(levels(modeldata$timedummy))!=1) { #modeldata$timedummy <- as.numeric(0)   
            model <- lm(expl_vars, data=modeldata, na.action = na.exclude)
            newdata <- modeldata
            newdata$timedummy <- newdata$timedummy[1]
            modeldata <- cbind(modeldata,fitted=predict.lm(model,newdata=newdata))

            modeldata$id <- 0
            for(i in 1:nrow(modeldata)) {
                teller <- teller + 1
                if(modeldata$id[i]==0) {
                    modeldata$id[i] <- teller
                    medium <- modeldata$med_code[i]
                    modeldata$distance <- abs(modeldata$fitted[i]-modeldata$fitted)/modeldata$fitted[i]
                    if(threshold=="nearest"){
                        modeldata$id[(modeldata$distance==min(modeldata$distance,na.rm=TRUE) & modeldata[,"id"]==0 & modeldata[,"med_code"]==medium)] <- teller
                    } else {
                        modeldata$id[(modeldata$distance<threshold & modeldata[,"id"]==0 & modeldata[,"med_code"]==medium)] <- teller
                    } 
                }
            }
        repdata <- rbind(repdata,modeldata)
        }
    }

    fullrep <- cbind(repsaledata(repdata$lnprice,repdata$counter,repdata$id),
                 repsaledata(repdata$lnarea,repdata$counter,repdata$id)[,4:5],
                 repsaledata(repdata$med_code,repdata$counter,repdata$id)[,4:5],
                 repsaledata(repdata$ah_code,repdata$counter,repdata$id)[,4:5],
                 repsaledata(repdata$lnsculpt_area,repdata$counter,repdata$id)[,4:5],
                 repsaledata(repdata$dum_signed,repdata$counter,repdata$id)[,4:5],
                 repsaledata(repdata$dum_dated,repdata$counter,repdata$id)[,4:5],
                 repsaledata(repdata$nr_works,repdata$counter,repdata$id)[,4:5])

    colnames(fullrep) <- c("id","time0","time1","price0","price1","area0","area1","med_code0","med_code1",
                        "ah_code0","ah_code1","sculpt0","sculpt1","sign0","sign1","date0","date1",
                        "nr0","nr1")

    repeatsales <- repsale(fullrep$price0,fullrep$time0,fullrep$price1,fullrep$time1,mergefirst=2,
                            graph=TRUE,graph.conf=TRUE,conf=.95)
    sps.RS_index <- exp(as.data.frame(repeatsales$pindex))*100

    dy <- fullrep$price1 - fullrep$price0
    timevar <- levels(factor(c(fullrep$time0, fullrep$time1)))
    nt = length(timevar)
    n = length(dy)
    xmat <- array(0, dim = c(n, nt - 1))
    for (j in seq(1 + 1, nt)) {
       xmat[, j - 1] <- ifelse(fullrep$time1 == timevar[j], 1, xmat[, j - 1])
       xmat[, j - 1] <- ifelse(fullrep$time0 == timevar[j],-1, xmat[, j - 1])
    }
    colnames(xmat) <- paste("Time", seq(1 + 1, nt))
    fit <- lm(dy ~ xmat + 0)

    fullrep$med_code0[is.na(fullrep$med_code0)] <- "Oil"
    fullrep$med_code1[is.na(fullrep$med_code1)] <- "Oil"
    fullrep$ah_code0[is.na(fullrep$ah_code0)] <- "Strauss"
    fullrep$ah_code1[is.na(fullrep$ah_code1)] <- "Strauss"

    darea <- fullrep$area1 - fullrep$area0

    med0 <- model.matrix(~fullrep$med_code0)
    med1 <- model.matrix(~fullrep$med_code1)
    dmed <- med1 - med0

    ah0 <- model.matrix(~fullrep$ah_code0)
    ah1 <- model.matrix(~fullrep$ah_code1)
    dah <- ah1 - ah0

    dsculpt <- fullrep$sculpt1 - fullrep$sculpt0
    dnr <- fullrep$nr1 - fullrep$nr0

    sign0 <- model.matrix(~fullrep$sign0)
    sign1 <- model.matrix(~fullrep$sign1)
    dsign <- sign1 - sign0

    date0 <- model.matrix(~fullrep$date0)
    date1 <- model.matrix(~fullrep$date1)
    ddate <- date1 - date0

    #ps.RS <- lm(dy ~ darea + dmed + dah + dsculpt + dnr + dsign + ddate + xmat + 0)
    ps.RS <- lm(dy ~ darea + dah + dsculpt + dnr + dsign + ddate + xmat + 0)
    ps.RS_results <- summary(ps.RS)$coefficients[grepl("Time", rownames(summary(ps.RS)$coefficients)),1]
    ps.RS_results <- as.data.frame(ps.RS_results)
    ps.RS_results$index_all <- exp(ps.RS_results$ps.RS_results)*100
    if(threshold=="nearest"){
        ps.RS_results <- cbind(ps.RS_results,sps.RS_index[c(3:62),])
    } else {
        ps.RS_results <- cbind(ps.RS_results,sps.RS_index[2:62,])
    } 
    ps.RS_results$pairs <- nrow(fullrep)
    
    return(ps.RS_results)
}

ps.RS_1 <- ps.RS(0.001)
ps.RS_2 <- ps.RS(0.0001)
ps.RS_n <- ps.RS("nearest")

rep_indices <- ps.RS_1[,c(2,3)] #time_results
rep_indices$Date <- c("2000Q2","2000Q3","2000Q4","2001Q1","2001Q2","2001Q3","2001Q4","2002Q1","2002Q2","2002Q3","2002Q4",
                      "2003Q1","2003Q2","2003Q3","2003Q4","2004Q1","2004Q2","2004Q3","2004Q4","2005Q1","2005Q2","2005Q3","2005Q4",
                      "2006Q1","2006Q2","2006Q3","2006Q4","2007Q1","2007Q2","2007Q3","2007Q4","2008Q1","2008Q2","2008Q3","2008Q4",
                      "2009Q1","2009Q2","2009Q3","2009Q4","2010Q1","2010Q2","2010Q3","2010Q4","2011Q1","2011Q2","2011Q3","2011Q4",
                      "2012Q1","2012Q2","2012Q3","2012Q4","2013Q1","2013Q2","2013Q3","2013Q4","2014Q1","2014Q2","2014Q3","2014Q4",
                      "2015Q1","2015Q2")

repeatsales_index$Date <- c("2000Q4","2001Q1","2001Q2","2001Q3","2001Q4","2002Q1","2002Q2","2002Q4",
                            "2003Q1","2003Q2","2003Q4","2004Q1","2004Q2","2004Q4","2005Q1","2005Q2","2005Q3","2005Q4",
                            "2006Q1","2006Q2","2006Q3","2006Q4","2007Q2","2007Q3","2007Q4","2008Q1","2008Q2","2008Q3","2008Q4",
                            "2009Q1","2009Q2","2009Q3","2009Q4","2010Q1","2010Q2","2010Q3","2010Q4","2011Q1","2011Q2","2011Q3","2011Q4",
                            "2012Q1","2012Q2","2012Q3","2012Q4","2013Q1","2013Q2","2013Q3","2013Q4","2014Q1","2014Q2","2014Q3","2014Q4",
                            "2015Q1","2015Q2")
rep_indices <- merge(rep_indices, repeatsales_index, by="Date", all=TRUE)

rep_indices <- cbind(rep_indices,Full_ps.RS2=ps.RS_2[,2])

ps.RS_n$Date <- c("2000Q3","2000Q4","2001Q1","2001Q2","2001Q3","2001Q4","2002Q1","2002Q2","2002Q3","2002Q4",
                  "2003Q1","2003Q2","2003Q3","2003Q4","2004Q1","2004Q2","2004Q3","2004Q4","2005Q1","2005Q2","2005Q3","2005Q4",
                  "2006Q1","2006Q2","2006Q3","2006Q4","2007Q1","2007Q2","2007Q3","2007Q4","2008Q1","2008Q2","2008Q3","2008Q4",
                  "2009Q1","2009Q2","2009Q3","2009Q4","2010Q1","2010Q2","2010Q3","2010Q4","2011Q1","2011Q2","2011Q3","2011Q4",
                  "2012Q1","2012Q2","2012Q3","2012Q4","2013Q1","2013Q2","2013Q3","2013Q4","2014Q1","2014Q2","2014Q3","2014Q4",
                  "2015Q1","2015Q2")
rep_indices <- merge(rep_indices, ps.RS_n, by="Date", all=TRUE)

rep_indices <- rep_indices[,c(-3,-6,-8,-9)]
colnames(rep_indices) <- c("Date","ps.RS(0.1%)","Repeat Sales","ps.RS(0.01%)","ps.RS(nearest)")    
```

```{r figure4, echo=FALSE, cache = TRUE, fig.height=4, fig.width=6, fig.cap="Pseudo repeat sales South African art price indices (2000Q1=100)"}
index_plot <- melt(rep_indices, id="Date")  # convert to long format
g <- ggplot(data=index_plot,aes(x=Date, y=value, group=variable, colour=variable)) 
g <- g + geom_point(size = 3) 
g <- g + geom_line()
g <- g + ylab("Index")
g <- g + xlab("")
g <- g + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
g <- g + theme(legend.title=element_blank()) + theme(legend.position="bottom")
g
```



##Evaluation

$M_t=M_{t-1}+r_t$ and $I_t=M_t+\epsilon_t=\sum_{i=1}^tr_i+\epsilon_t$

$r_t^*=I_t-I_{t-1}=r_t+(\epsilon_t-\epsilon_{t-1})= r_t+\eta_t$

$Vol=\sigma_{r_t^*}=\sqrt{\sigma_r^2+\sigma_\eta^2}$ and $AC(1)=\rho_{r^*}=(\rho_r\sigma_r^2-\sigma_\eta^2/2)/(\sigma_r^2+\sigma_\eta^2)$ 



#Bubble Detection

$$\Delta y_t = \alpha_w + (\rho_w-1)y_{t-1}+\sum_{i=1}^k\phi_w^i \Delta y_{t-i} + \epsilon_t $$


#References
